### 2015-03-25 swmpr_manu.Rnw
finished metabolism results and applications section

% Template for PLoS
% Version 3.1 February 2015
%
% To compile to pdf, run:
% latex plos.template
% bibtex plos.template
% latex plos.template
% latex plos.template
% dvipdf plos.template
%
% % % % % % % % % % % % % % % % % % % % % %
%
% -- IMPORTANT NOTE
%
% This template contains comments intended 
% to minimize problems and delays during our production 
% process. Please follow the template instructions
% whenever possible.
%
% % % % % % % % % % % % % % % % % % % % % % % 
%
% Once your paper is accepted for publication, 
% PLEASE REMOVE ALL TRACKED CHANGES in this file and leave only
% the final text of your manuscript.
%
% There are no restrictions on package use within the LaTeX files except that 
% no packages listed in the template may be deleted.
%
% Please do not include colors or graphics in the text.
%
% Please do not create a heading level below \subsection. For 3rd level headings, use \paragraph{}.
%
% % % % % % % % % % % % % % % % % % % % % % %
%
% -- FIGURES AND TABLES
%
% Please include tables/figure captions directly after the paragraph where they are first cited in the text.
%
% DO NOT INCLUDE GRAPHICS IN YOUR MANUSCRIPT
% - Figures should be uploaded separately from your manuscript file. 
% - Figures generated using LaTeX should be extracted and removed from the PDF before submission. 
% - Figures containing multiple panels/subfigures must be combined into one image file before submission.
% For figure citations, please use "Fig." instead of "Figure".
% See http://www.plosone.org/static/figureGuidelines for PLOS figure guidelines.
%
% Tables should be cell-based and may not contain:
% - tabs/spacing/line breaks within cells to alter layout or alignment
% - vertically-merged cells (no tabular environments within tabular environments, do not use \multirow)
% - colors, shading, or graphic objects
% See http://www.plosone.org/static/figureGuidelines#tables for table guidelines.
%
% For tables that exceed the width of the text column, use the adjustwidth environment as illustrated in the example table in text below.
%
% % % % % % % % % % % % % % % % % % % % % % % %
%
% -- EQUATIONS, MATH SYMBOLS, SUBSCRIPTS, AND SUPERSCRIPTS
%
% IMPORTANT
% Below are a few tips to help format your equations and other special characters according to our specifications. For more tips to help reduce the possibility of formatting errors during conversion, please see our LaTeX guidelines at http://www.plosone.org/static/latexGuidelines
%
% Please be sure to include all portions of an equation in the math environment.
%
% Do not include text that is not math in the math environment. For example, CO2 will be CO\textsubscript{2}.
%
% Please add line breaks to long display equations when possible in order to fit size of the column. 
%
% For inline equations, please do not include punctuation (commas, etc) within the math environment unless this is part of the equation.
%
% % % % % % % % % % % % % % % % % % % % % % % % 
%
% Please contact latex@plos.org with any questions.
%
% % % % % % % % % % % % % % % % % % % % % % % %

@@

The \texttt{qaqc} function is a simple screen to retain values from the data with specified QAQC flags (described \href{http://cdmo.baruch.sc.edu/data/qaqc.cfm}{here}).  Each parameter in the \texttt{swmpr} data typically has a corresponding QAQC column of the same name with the added prefix \texttt{f\_}.  Values in the QAQC column specify a flag from -5 to 5.  Generally, only data with the  \texttt{0} \ac{QAQC} flag should be used, which is the default option for the \texttt{QAQC} function.  Data that do not satisfy \ac{QAQC} criteria are converted to \texttt{NA} values.   Additionally, simple filters are used to remove obviously bad values, e.g., wind speed values less than zero or pH values greater than 12. Erroneous data entered as -99 are also removed. Processed data will have QAQC columns removed, in addition to removal of values in the actual parameter columns that do not meet the criteria.

@@

Viewing the number of observations for each parameter that are assigned to a \ac{QAQC} flag may be useful for deciding how to process the data with ), \texttt{qaqc}.  The \texttt{qaqcchk} function can be used to view this information.  Consult the \href{http://cdmo.baruch.sc.edu/data/qaqc.cfm}{online documentation} for a description of each QAQC flag.

@@

A subset method added to the existing \texttt{subset} function is available for \texttt{swmpr} objects.  This function is used to subset the data by date and/or a selected parameter.  The date can be a single value or as two dates to select records within the range. The former case requires a binary operator input as a character string passed to the argument, such as \texttt{>} or \texttt{<}.  The subset argument for the date(s) must also be a character string of the format YYYY-mm-dd HH:MM for each element (i.e., \%Y-\%m\%-\%d \%H:\%M in POSIX standards).  Be aware that an error may be returned using this function if the subset argument is in the correct format but the calendar date does not exist, e.g. \texttt{2012-11-31 12:00}.  Finally, the function can be used to remove rows and columns that do not contain data.

@@

The analysis functions range from general purpose tools for time series analysis to more specific functions for working with continuous monitoring data in estuaries (\Cref{tab:analyze}).  The latter category includes a limited number of functions that were developed by myself or others.  The general purpose tools are \texttt{swmpr} methods that were developed for existing generic functions in the R base installation or relevant packages.  These functions include \texttt{swmpr} methods for \texttt{aggregate}, \texttt{filter}, and \texttt{approx} to deal with missing or noisy data and more general functions for exploratory data analaysis, such as \texttt{plot}, \texttt{summary}, and \texttt{hist} methods.  Decomposition functions \texttt{decomp} and \texttt{decomp\_cj}) are provided as relatively simple approaches for decomposing time series into additive or multiplicative components. The analysis functions may or may not return a \texttt{swmpr} object depending on whether further processing with \texttt{swmpr} methods is possible from the output.

@@

Water quality and weather observations from January 1995 to December 2014 for all \ac{NERRS} sites were obtained through a bulk data request using the \href{http://cdmo.baruch.sc.edu/aqs/zips.cfm}{zip downloads} feature of \ac{CDMO}.  After the download was complete, all csv files for each station were imported into R using the \texttt{import\_local} function, processed using the \texttt{setstep} and \texttt{qaqc} functions, then saved on a local hard drive as binary RData files (see \href{https://github.com/fawda123/swmp_rats/blob/master/R/dat_proc.R}{here} for the R script). This resulted in a single \texttt{swmpr} object for each paramter at each site.  All files were then uploaded to a remote server.  An \href{https://gist.github.com/fawda123/4fc51c2cb86341ed9291}{R script} was executed that retrieved and processed combined the water quality and weather data for each site.

<<metab_plo, eval = T, cache = T, echo = F, fig.cap = 'Aggregated estimates of net metabolism, gross production, and total respiration for two sites at each \\ac{NERRS} reserve.  Values are daily integrated estimates as mean annual values averaged across all years with 95\\% confidence intervals.  Two sites were chosen from each reserve that had the longest available time series.', fig.height = 6, fig.width = 8>>=
# # packages to use
# library(SWMPr)
# library(httr)
# library(XML)
# library(foreach)
# library(doParallel)
# 
# # names of files on server
# files_s3 <- httr::GET('https://s3.amazonaws.com/swmpalldata/')$content
# files_s3 <- rawToChar(files_s3)
# files_s3 <- htmlTreeParse(files_s3, useInternalNodes = T)
# files_s3 <- xpathSApply(files_s3, '//contents//key', xmlValue)
# files_s3 <- gsub('\\.RData$', '', files_s3) 
# 
# # find only active water quality and weather sites
# # IP address must be registed with CDMO 
# meta <- site_codes()
# sel <- meta$status %in% 'Active' & !grepl('nut$', meta$station_code)
# meta <- meta[sel, ]
# 
# # get wq sites from meta, then filter by those on server
# wq_sites <- grep('wq$', meta$station_code, value = T)
# wq_sites <- files_s3[gsub('\\.RData$', '', files_s3) %in% wq_sites]
# 
# # setup parallel backend for processing
# cl <- makeCluster(8)
# registerDoParallel(cl)
# strt <- Sys.time()
# 
# # process
# metabs <- foreach(wq_site = wq_sites) %dopar% {
#  
#   library(SWMPr)
#   
#   # progress
#   sink('log.txt')
#   cat(wq_site, which(wq_site == wq_sites), 'of', length(wq_sites), '\n')
#   print(Sys.time() - strt)
#   sink()
#   
#   # find corresponding wx station
#   met_site <- substr(wq_site, 1, 3)
#   met_site <- grep(paste0('^', met_site, '.*met'), files_s3, value = T)
#   met_site <- gsub('\\.RData$', '', met_site)
# 
#   # continue if wx data found
#   if(length(met_site) > 0){
#     
#     # if > 1 wx site, pick the one with more obs
#     if(length(met_site) > 1){
# 
#       met_list <- vector('list', length(met_site))
#       names(met_list) <- met_site
#       for(met in met_site){
#       
#         met_tmp <- import_remote(met)
#         met_list[[met]] <- met_tmp
#       
#       }
#     
#       # pick the one with most obs
#       met_most <- which.max(unlist(lapply(met_list, nrow)))
#       met <- met_list[met_most][[1]]
# 
#     # otherwise load only one
#     } else {
#     
#       # met
#       met <- import_remote(met_site)
#     
#     }
#     
#     ##
#     # load the wq file
#     
#     # wq
#     wq <- import_remote(wq_site)
#     
#     ## 
#     # combine, estimate metabolism, reduce data volume
#     dat <- comb(wq, met, method = attr(met, 'station'))
#     dat <- ecometab(dat)
#     dat <- dat[1, ]
#     
#     dat
#     
#   }
#   
# }
# names(metabs) <- wq_sites
# save(metabs, file = 'metabs.RData')

@@

@@

#reassign factors for ranking in plot
sort_val <- order(to_plo[to_plo$metab == 'Pg', 'value'])
to_plo$site <- factor(gsub('wq$', '', to_plo$site))
to_plo$site <- factor(
  to_plo$site,
  levels = levels(to_plo$site)[sort_val],
  labels = levels(to_plo$site)[sort_val]
  )
new_labs <- c('Mean net annual metabolism', 'Mean annual gross production', 'Mean annual total respiration')
to_plo$metab <- factor(to_plo$metab, levels = c('NEM', 'Pg', 'Rt'), labels = new_labs)

ylabs <- expression(paste('g ', O[2],' ', m^-2,' ', d^-1))
p <- ggplot(to_plo, 
    aes(x = site, y = value, group = metab, fill = metab)
  ) + 
  geom_bar(stat = 'identity', width = 1, colour = 'black') + 
  scale_y_continuous(ylabs) + 
  facet_wrap( ~ metab, ncol = 1, scales = 'free_y') + 
  scale_x_discrete(name = element_blank()) + 
  theme_classic() + 
  theme(
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, 
      size = 8),
    legend.position = 'none'
    ) +
  geom_errorbar(aes(ymin = lower, ymax = upper), width = 0)
print(p)

@@

% table summary of trends
<<eval = F, cache = T, echo = F>>=

@@

# summarize for annual means, get conf ints
metabs_tab <- gather(metabs_sum, 'metab', 'value', 4:6)
metabs_tab$yr_grp <- '1995 - 2004'
metabs_tab$yr_grp[metabs_tab$year > 2004] <- '2005-2014'
metabs_tab <- mutate(metabs_tab, value = 0.032 * value) %>% 
  group_by(metabs_tab, site, metab, yr_grp)

# summarize trends

@@

The second and third examples are two Shiny web applications \cite{Chang15} that illustrate summaries and comparisons of \ac{SWMP} data within and between reserves (see the \nameref{supp_info}).  The second application evaluates trends in \ac{SWMP} data within and between sites using an interactive map.  Trends between reserves can be viewed using the map, whereas trends at individual sites can be viewed by clicking on a map location.  Site-levels trends are described below the map using a simple linear regression to show an increase or decrease in values over time.  Trends on the map at each station are plotted as circles that identify the direction and significance of the trend.  The trend direction is blue for decreasing and red for increasing.  The significance is indicated by radius of the circle and color shading where larger points with darker colors indicate a strong trend.  The third application provides graphical summaries of water quality, weather, or nutrient station data at individual stations. The drop down menus can be used to select the station, date range, and parameter for plotting. Plots are based on daily averages for each parameter.  The data used for each application are similar to those used to estimate ecosystem metabolism described above.

>>>>

% Template for PLoS
% Version 3.1 February 2015
%
% % % % % % % % % % % % % % % % % % % % % % % 
%
% Once your paper is accepted for publication, 
% PLEASE REMOVE ALL TRACKED CHANGES in this file and leave only
% the final text of your manuscript.
%
% There are no restrictions on package use within the LaTeX files except that 
% no packages listed in the template may be deleted.
%
% Please do not include colors or graphics in the text.
%
% Please do not create a heading level below \subsection. For 3rd level headings, use \paragraph{}.
%
% % % % % % % % % % % % % % % % % % % % % % %
%
% -- FIGURES AND TABLES
%
% Please include tables/figure captions directly after the paragraph where they are first cited in the text.
%
% DO NOT INCLUDE GRAPHICS IN YOUR MANUSCRIPT
% - Figures should be uploaded separately from your manuscript file. 
% - Figures generated using LaTeX should be extracted and removed from the PDF before submission. 
% - Figures containing multiple panels/subfigures must be combined into one image file before submission.
% For figure citations, please use "Fig." instead of "Figure".
% See http://www.plosone.org/static/figureGuidelines for PLOS figure guidelines.
%
% Tables should be cell-based and may not contain:
% - tabs/spacing/line breaks within cells to alter layout or alignment
% - vertically-merged cells (no tabular environments within tabular environments, do not use \multirow)
% - colors, shading, or graphic objects
% See http://www.plosone.org/static/figureGuidelines#tables for table guidelines.
%
% For tables that exceed the width of the text column, use the adjustwidth environment as illustrated in the example table in text below.
%
% % % % % % % % % % % % % % % % % % % % % % % %
%
% -- EQUATIONS, MATH SYMBOLS, SUBSCRIPTS, AND SUPERSCRIPTS
%
% IMPORTANT
% Below are a few tips to help format your equations and other special characters according to our specifications. For more tips to help reduce the possibility of formatting errors during conversion, please see our LaTeX guidelines at http://www.plosone.org/static/latexGuidelines
%
% Please be sure to include all portions of an equation in the math environment.
%
% Do not include text that is not math in the math environment. For example, CO2 will be CO\textsubscript{2}.
%
% Please add line breaks to long display equations when possible in order to fit size of the column. 
%
% For inline equations, please do not include punctuation (commas, etc) within the math environment unless this is part of the equation.
%
% % % % % % % % % % % % % % % % % % % % % % % %

@@

The \texttt{qaqc} function is a simple screen to retain values from the data with specified QAQC flags (see the \ac{CDMO} descriptions \href{http://cdmo.baruch.sc.edu/data/qaqc.cfm}{here}).  Each parameter in the \texttt{swmpr} data typically has a corresponding QAQC column of the same name with the added prefix \texttt{f\_}.  Values in the QAQC column specify a flag from -5 to 5.  Generally, only data with the  \texttt{0} \ac{QAQC} flag should be used, which is the default option for the \texttt{QAQC} function.  Data that do not satisfy \ac{QAQC} criteria are converted to \texttt{NA} values.   Additionally, simple filters are used to remove obviously bad values, e.g., wind speed values less than zero or pH values greater than 12. Erroneous data entered as -99 are also removed. Processed data will have QAQC columns removed, in addition to removal of values in the actual parameter columns that do not meet the criteria.

@@

Viewing the number of observations for each parameter that are assigned to a \ac{QAQC} flag may be useful for deciding how to process the data with \texttt{qaqc}.  The \texttt{qaqcchk} function can be used to view this information.

@@

A subset method added to the existing generic \texttt{subset} function in R is available for \texttt{swmpr} objects.  This function is used to subset the data by date and/or a selected parameter.  The date can be a single value or as two dates to select records within the range. The former case requires a binary operator input as a character string passed to the argument, such as \texttt{>} or \texttt{<}.  The subset argument for the date(s) must also be a character string of the format YYYY-mm-dd HH:MM for each element (i.e., \%Y-\%m\%-\%d \%H:\%M in POSIX standards).  Be aware that an error may be returned using this function if the subset argument is in the correct format but the calendar date does not exist, e.g. \texttt{2012-11-31 12:00}.  Finally, the function can be used to remove rows and columns that do not contain data.

@@

The analysis functions range from general purpose tools for time series analysis to more specific functions for working with continuous monitoring data in estuaries (\Cref{tab:analyze}).  The latter category includes a limited number of functions that were developed by myself or others.  The general purpose tools are \texttt{swmpr} methods that were developed for existing generic functions in the R base installation or relevant packages.  These functions include \texttt{swmpr} methods for \texttt{aggregate}, \texttt{filter}, and \texttt{approx} to deal with missing or noisy data and more general functions for exploratory data analaysis, such as \texttt{plot}, \texttt{summary}, and \texttt{hist} methods.  Decomposition functions, \texttt{decomp} and \texttt{decomp\_cj}, are provided as relatively simple approaches for decomposing time series into additive or multiplicative components. The analysis functions may or may not return a \texttt{swmpr} object depending on whether further processing with \texttt{swmpr} methods is possible from the output.

@@

Water quality and weather observations from January 1995 to December 2014 for all \ac{NERRS} sites were obtained through a bulk data request using the \href{http://cdmo.baruch.sc.edu/aqs/zips.cfm}{zip downloads} feature of \ac{CDMO}.  All csv files for each station were imported into R using the \texttt{import\_local} function, processed using the \texttt{setstep} and \texttt{qaqc} functions, then saved on a local hard drive as binary RData files (see \href{https://github.com/fawda123/swmp_rats/blob/master/R/dat_proc.R}{here} for the R script). This resulted in a single \texttt{swmpr} object for each parameter at each site.  All files were then uploaded to a remote server.  An additional \href{https://gist.github.com/fawda123/4fc51c2cb86341ed9291}{R script} was executed that retrieved and processed combined the water quality and weather data for each site to esimate metabolism.  For brevity, only two sites at each reserve were evaluated.  Mean annual values at each site, organized by region, are shown in \Cref{fig:metab_plo}, whereas decadal comparisons are shown in \Cref{tab:metab_tab}.  All sites were generally net heterotrophic across the range of observations (i.e., sink of organic matter, in agreement with Caffrey \cite{Caffrey03}), although differences were observed in early (i.e., 1995-2004) as compared to recent (2005-2014) time periods. Overall, the results indicate that between-region and within-site differences in metabolism are apparent and varying by time period, such that a more comprehensive evaluation of factors that influence metabolic rates is needed.  More importantly, the use of the data retreival, synthesis, and analysis functions to create the results shown in \Cref{fig:metab_plo,tab:metab_tab} illustrates the utility provided by the SWMPr package.

% metabolism figure
<<metab_plo, eval = T, cache = T, echo = F, fig.cap = 'Aggregated estimates of net metabolism, gross production, and total respiration for two sites at each \\ac{NERRS} reserve.  Values are daily integrated estimates as mean annual values averaged across all years with 95\\% confidence intervals.  Two sites were chosen from each reserve that had the longest available time series. Regions are approximations of geographic coordinates and are not exact, e.g., Old Woman Creek (Lake Erie) is Atlantic.', fig.height = 6, fig.width = 8, out.width = '\\textwidth'>>=
# # packages to use
# library(SWMPr)
# library(httr)
# library(XML)
# library(foreach)
# library(doParallel)
# 
# # names of files on server
# files_s3 <- httr::GET('https://s3.amazonaws.com/swmpalldata/')$content
# files_s3 <- rawToChar(files_s3)
# files_s3 <- htmlTreeParse(files_s3, useInternalNodes = T)
# files_s3 <- xpathSApply(files_s3, '//contents//key', xmlValue)
# files_s3 <- gsub('\\.RData$', '', files_s3) 
# 
# # find only active water quality and weather sites
# # IP address must be registed with CDMO 
# meta <- site_codes()
# sel <- meta$status %in% 'Active' & !grepl('nut$', meta$station_code)
# meta <- meta[sel, ]
# 
# # get wq sites from meta, then filter by those on server
# wq_sites <- grep('wq$', meta$station_code, value = T)
# wq_sites <- files_s3[gsub('\\.RData$', '', files_s3) %in% wq_sites]
# 
# # setup parallel backend for processing
# cl <- makeCluster(8)
# registerDoParallel(cl)
# strt <- Sys.time()
# 
# # process
# metabs <- foreach(wq_site = wq_sites) %dopar% {
#  
#   library(SWMPr)
#   
#   # progress
#   sink('log.txt')
#   cat(wq_site, which(wq_site == wq_sites), 'of', length(wq_sites), '\n')
#   print(Sys.time() - strt)
#   sink()
#   
#   # find corresponding wx station
#   met_site <- substr(wq_site, 1, 3)
#   met_site <- grep(paste0('^', met_site, '.*met'), files_s3, value = T)
#   met_site <- gsub('\\.RData$', '', met_site)
# 
#   # continue if wx data found
#   if(length(met_site) > 0){
#     
#     # if > 1 wx site, pick the one with more obs
#     if(length(met_site) > 1){
# 
#       met_list <- vector('list', length(met_site))
#       names(met_list) <- met_site
#       for(met in met_site){
#       
#         met_tmp <- import_remote(met)
#         met_list[[met]] <- met_tmp
#       
#       }
#     
#       # pick the one with most obs
#       met_most <- which.max(unlist(lapply(met_list, nrow)))
#       met <- met_list[met_most][[1]]
# 
#     # otherwise load only one
#     } else {
#     
#       # met
#       met <- import_remote(met_site)
#     
#     }
#     
#     ##
#     # load the wq file
#     
#     # wq
#     wq <- import_remote(wq_site)
#     
#     ## 
#     # combine, estimate metabolism, reduce data volume
#     dat <- comb(wq, met, method = attr(met, 'station'))
#     dat <- ecometab(dat)
#     dat <- dat[1, ]
#     
#     dat
#     
#   }
#   
# }
# names(metabs) <- wq_sites
# save(metabs, file = 'metabs.RData')

@@

# get region
data(stat_locs)
names(stat_locs)[names(stat_locs) %in% 'station_code'] <- 'site'
metabs_plo$site <- gsub('wq$', '', metabs_plo$site)
metabs_plo <- left_join(metabs_plo, stat_locs, by = 'site')
region <- rep('Pacific', nrow(metabs_plo))
region[metabs_plo$longitude > -110] <- 'Gulf'
region[metabs_plo$latitude > 28 & metabs_plo$longitude > -84] <- 'Atlantic'

# combine, select, sort
metabs_plo <- select(metabs_plo, -c(latitude, longitude, gmt_off))
metabs_plo$region <- region
metabs_plo <- arrange(metabs_plo, region, site)

@@

#reassign factors for ranking in plot
sort_val <- to_plo[to_plo$metab == 'Pg', c('region', 'value', 'site')]
sort_val <- order(sort_val$region, sort_val$value)
to_plo$site <- factor(to_plo$site)
to_plo$site <- factor(
  to_plo$site,
  levels = levels(to_plo$site)[sort_val],
  labels = levels(to_plo$site)[sort_val]
  )
new_labs <- c('Mean net metabolism', 'Mean gross production', 'Mean total respiration')
to_plo$metab <- factor(to_plo$metab, levels = c('NEM', 'Pg', 'Rt'), labels = new_labs)

ylabs <- expression(paste('g ', O[2],' ', m^-2,' ', d^-1))
library(ggplot2)
p <- ggplot(to_plo, 
    aes(x = site, y = value, group = metab, fill = metab)
  ) + 
  geom_bar(stat = 'identity', width = 1, colour = 'black') + 
  scale_y_continuous(ylabs) + 
  facet_grid(metab ~ region, scales = 'free') + 
  scale_x_discrete(name = element_blank()) + 
  theme_bw() + 
  theme(
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, 
      size = 7),
    legend.position = 'none'
    ) +
  geom_errorbar(aes(ymin = lower, ymax = upper), width = 0)
print(p)

@@

% table summary of trends
<<eval = T, results = 'asis', cache = T, echo = F>>=

@@

# summarize by time periods
metabs_tab <- gather(metabs_sum, 'metab', 'value', 4:6)
metabs_tab$yr_grp <- '1995-2004'
metabs_tab$yr_grp[metabs_tab$date > 2004] <- '2005-2014'
metabs_tab <- mutate(metabs_tab, value = 0.032 * value) %>% 
  group_by(site, metab, yr_grp) %>% 
  arrange(yr_grp, date, metab) %>% 
  summarize(ave = round(mean(value, na.rm = TRUE), 2)) %>% 
  spread(yr_grp, ave) 
trends <- apply(metabs_tab, 1, function(x){

  val <- as.numeric(x[['2005-2014']])
  trnd <- val > as.numeric(x[['1995-2004']])

  if(is.na(trnd)) return(val)
  if(trnd) return(paste0('{\\bf ', val, '}'))
  if(!trnd) return(paste0('{\\it ', val, '}'))

})
metabs_tab[, '2005-2014'] <- trends
metabs_tab <- gather(metabs_tab, 'period', 'value', 3:4) %>% 
  unite('var', metab, period, sep = ' ') %>% 
  spread(var, value) %>% 
  mutate(site = gsub('wq$', '', site))

# get region
data(stat_locs)
names(stat_locs)[names(stat_locs) %in% 'station_code'] <- 'site'
metabs_tab <- left_join(metabs_tab, stat_locs, by = 'site')
region <- rep('Pacific', nrow(metabs_tab))
region[metabs_tab$longitude > -110] <- 'Gulf'
region[metabs_tab$latitude > 28 & metabs_tab$longitude > -84] <- 'Atlantic'

# combine, select, sort
metabs_tab <- select(metabs_tab, -c(latitude, longitude, gmt_off)) %>% 
  mutate(region = region) %>% 
  arrange(region, site)

nrgroups <- as.numeric(table(metabs_tab$region))
# table

cap <- 'Trends in metabolism for two sites at each of the \\ac{NERRS} reserves.  Values are averages of mean annual estimates for each period of observation (1994-2004 and 2005-2014).  Bold values indicate an increase from the first period, whereas italic values indicate a decrease.  Regions are approximations of geographic coordinates and are not exact, e.g., Old Woman Creek (Lake Erie) is Atlantic.' 
foots <- '\\footnotesize{\\textsuperscript{{\\it a}}NEM: net ecosystem metabolism, Pg: gross production, Rt: total respiration, all values in g O$_2$ m$^{-2}$ d$^{-1}$ as annual averages.}'

library(Hmisc)
latex(
  metabs_tab[, !names(metabs_tab) %in% c('site', 'region')], 
  file = '', 
  caption = cap, 
  cgroup = c('NEM\\textsuperscript{{\\it a}}', 'Pg', 'Rt'), 
  n.cgroup = c(2, 2, 2), 
  colheads = rep(c('1995-2004', '2005-2014'), 3),
  rowlabel = 'Site', 
  rowname = metabs_tab$site,
  digits = 2,
  insert.bottom = foots,
  rgroup = unique(metabs_tab$region),
  n.rgroup = nrgroups,
  label = 'tab:metab_tab',
  size = 'footnotesize'
  )

@@

The second and third examples are two Shiny web applications \cite{Chang15} that illustrate summaries and comparisons of \ac{SWMP} data within and between reserves that were created using the SWMPr package (see the \nameref{supp_info}).  The second application evaluates trends in \ac{SWMP} data within and between sites using an interactive map.  Trends between reserves can be viewed using the map, whereas trends at individual sites can be viewed by clicking on a map location.  Site-levels trends are described below the map using a simple linear regression to show an increase or decrease in values over time.  Trends on the map at each station are plotted as circles that identify the direction and significance of the trend.  The trend direction is blue for decreasing and red for increasing.  The significance is indicated by radius of the circle and color shading where larger points with darker colors indicate a strong trend.  The third application provides graphical summaries of water quality, weather, or nutrient station data at individual stations. The drop down menus can be used to select the station, date range, and parameter for plotting. Plots are based on daily averages for each parameter.  The data used for each application are similar to those used to estimate ecosystem metabolism described above.
